Towards atomistic resolution structure of phosphatidylcholine headgroup and glycerol backbone at different ambient conditions\footnote{Publication about results presented in the NMRlipids project}

### 2015-10-11 HGmodelMANUSCRIPT/REPLYtoREVIEWER2.txt

Reviewer 2.

We thank for the very positive feedback. The errors in english are
now fixed as suggested.

Reviewer1.

1.
Author reply:

2.
Author reply:
The subjective sweet spots span +-0.02 AROUND WHAT? IS THIS THE AVERAGE
OF EXPERIMENTAL VALUE?

3.
Author reply:
The title of our manuscript is:
"Towards atomistic resolution structure of
phosphatidylcholine headgroup and glycerol
backbone at different ambient conditions"
in the introduction we write:
"Phospholipid molecules are composed of hydrophobic acyl chains connected
by a glycerol backbone to a hydrophilic headgroup ..."
This is widely used naming convention for phospholipid molecules.
According to the title and the naming convetion used in the manuscript, 
the beginning of acyl chains are not the part of headgroup and are 
beoynd the scope defined in the title. Their presence at the interface
or their chemical environment do not change this.

We are not willing to extend the scope of the manuscript since it
would signifincantly increase already massive information content and 
length. Instead, the acyl chains order parameters will be discussed in 
the context of the NMRLipids III project, see 
http://nmrlipids.blogspot.fi/p/about.html.

### 2015-10-11 HGmodelMANUSCRIPT/HGmodel_ACStemplate.tex
Add TOC in the manuscript
English corrections by the reviewer and typo from https://github.com/NMRLipids/nmrlipids.blogspot.fi/issues/65 were fixed. The first draft of the second reply to the reviewers added

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The "tocentry" environment can be used to create an entry for the
%% graphical table of contents. It is given here as some journals
%% require that it is printed as part of the abstract page. It will
%% be automatically moved as appropriate.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\begin{tocentry}

%Some journals require a graphical entry for the Table of Contents.
%This should be laid out ``print ready'' so that the sizing of the
%text is correct.

@@

%\end{tocentry}

@@

\subsection{Order parameters from experiments}\label{expORDp}
The order parameter of a hydrocarbon C--H vector is defined as 
\begin{equation}\label{orderP}
S_{{\rm CH}}=\frac{1}{2}\langle 3 \cos^2 \theta-1 \rangle,
\end{equation} 
where the angle brackets denote an ensemble average over the sampled conformations, and $\theta$ is the angle between the C--H bond and the membrane normal.
The absolute values of order parameters can be measured by detecting quadrupolar splitting with $^2$H NMR~\cite{seelig77c} or by detecting dipolar 
splitting with $^1$H-$^{13}$C NMR~\cite{hong95a,gross97,dvinskikh05a,ferreira13}. The measurements are based on
different physical interactions, and also the connections between order parameter and quadrupolar or dipolar splitting
are different. The absolute values of order parameters from the measured quadrupolar splitting $\Delta \nu_Q$ ($^2$H NMR) are calculated using 
the equation $|S_{{\rm CD}}|=\frac{4}{3} \frac{e^2qQ}{h} \Delta \nu_Q$, where the value for the static quadrupole
splitting constant is estimated from various experiments to be 170~kHz leading to a numerical relation $|S_{{\rm CD}}|=0.00784 \times \Delta \nu_Q$~\cite{seelig77c}. 
The absolute values of order parameters from the effective dipolar coupling $d_\mathrm{CH}$ ($^1$H-$^{13}$C NMR) are calculated using the equation
$|S_{{\rm CH}}|=\frac{4\pi\langle r_\mathrm{CH}^3 \rangle}{\hbar \mu_0 \gamma_h \gamma_c} d_\mathrm{CH}$, where
values between 20.2--22.7~kHz are used for $\frac{4\pi\langle r_\mathrm{CH}^3 \rangle}{\hbar \mu_0 \gamma_h \gamma_c}$,
depending on the original authors~\cite{hong95a,gross97,dvinskikh05a,ferreira13}.
The effective dipolar coupling $d_\mathrm{CH}$ is related to the measured dipolar splitting $\Delta \nu_\mathrm{CH}$ 
through a scaling factor that depends on the pulse sequence used in the $^1$H-$^{13}$C NMR experiment~\cite{hong95a,gross97,dvinskikh05a,ferreira13}.
It is important to note that the order parameters measured with different techniques based on different physical interactions are in good agreement
with each other (see Results and Discussion), indicating very high quantitative accuracy of the measurements.
For a more detailed discussion, see Ref.~\citenum{accuracyPOST}.

@@

To get a general idea of the quality of the glycerol backbone and choline headgroup structures in different models, we calculated 
the order parameters for these parts from thirteen different lipid models (Table~\ref{systems}) and 
plotted the results together with experimental values in Fig.~\ref{HGorderparameters}.
Two criteria were used to judge the quality of the model: there must not be significant  {\bf forking} in the $\alpha$ and $\beta$ carbons,
there must be only moderate forking in the g$_3$ carbon and there must be significant forking in the g$_1$ carbon, the {\bf magnitude}
should be preferably inside to the subjective sweet spots determined from experiments (blue shaded regions in Fig.~\ref{HGorderparameters}).
The results for each force field in respect to the above criteria are summarized in Figure~\ref{FullHydrationComparisonTable}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{../DATAreportediINblog/comparisonTable.pdf}
  \newline
  \caption{\label{FullHydrationComparisonTable}
Rough subjective ranking of force fields based on Fig.~\ref{HGorderparameters}.
%
Here "M" indicates a magnitude problem, "F" a forking problem;
letter size increases with problem severity.
%
Color scheme:
"within experimental error" (dark green),
"almost within experimental error" (light green),
"clear deviation from experiments" (light red), and
"major deviation from experiments" (dark red).
%
The $\Sigma$-column shows the total deviation of the force field,
when individual carbons are given weights of 0 (matches experiment), 1, 2, and 4 (major deviation).
%
For full details of the assessment, see Supporting Information.
} 
\end{figure}

None of the studied force fields fulfils these criteria completely, however CHARMM36 is pretty close. 
This is not surprising since the dihedral potentials in this model are tuned to reproduce these parameters better against experiments~\cite{klauda10}.
The next models in the list are CHARMM36-UA~\cite{henin08,lee14} and H\"ogberg08~\cite{hogberg08} which is also not surprising since
these models are using CHARMM bonded potentials for glycerol backbone and choline. The fourth and the fifth models in the list, MacRog~\cite{maciejewski14} and
GAFFlipid~\cite{dickson12}, have independently determined dihedral potentials. All the models based on Gromos potentials and Slipids perform less well.
In the present work we subject the CHARMM36, MacRog, GAFFlipid and Berger-POPC-07 to a more careful comparison including the sterospecific labeling  
(Fig.~\ref{HGorderparameters2}), and atomistic level structure and responses to the dehydration and cholesterol content in the following sections.
These models are selected for more detailed studies since they are the best representatives of different dihedral potential parametrization techniques 
(CHARMM36, MacRog, GAFFlipid), and the Berger based models are the most used lipid model in the literature.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The "tocentry" environment can be used to create an entry for the
%% graphical table of contents. It is given here as some journals
%% require that it is printed as part of the abstract page. It will
%% be automatically moved as appropriate.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{tocentry}

%\begin{figure}[]
%  \centering
  \includegraphics[width=9cm]{../Fig/TOCfinal.png}

%  \caption{\label{POPCstructure}
%    Chemical structure of  1-palmitoyl-2-oleoylphosphatidylcholine (POPC).}

%\end{figure}
%Some journals require a graphical entry for the Table of Contents.
%This should be laid out ``print ready'' so that the sizing of the
%text is correct.

@@

\end{tocentry}

@@

\subsection{Order parameters from experiments}\label{expORDp}
The order parameter of a hydrocarbon C--H vector is defined as 
\begin{equation}\label{orderP}
S_{{\rm CH}}=\frac{1}{2}\langle 3 \cos^2 \theta-1 \rangle,
\end{equation} 
where the angle brackets denote an ensemble average over the sampled conformations, and $\theta$ is the angle between the C--H bond and the membrane normal.
The absolute values of order parameters can be measured by detecting quadrupolar splitting with $^2$H NMR~\cite{seelig77c} or by detecting dipolar 
splitting with $^1$H-$^{13}$C NMR~\cite{hong95a,gross97,dvinskikh05a,ferreira13}. The measurements are based on
different physical interactions, and also the connections between order parameter and quadrupolar or dipolar splitting
are different. The absolute values of order parameters from the measured quadrupolar splitting $\Delta \nu_Q$ ($^2$H NMR) are calculated using 
the equation $|S_{{\rm CD}}|=\frac{4}{3} \frac{h}{e^2qQ} \Delta \nu_Q$, where the value for the static quadrupole
splitting constant is estimated from various experiments to be 170~kHz leading to a numerical relation 
$|S_{{\rm CD}}|=0.00784 \times \Delta \nu_Q$~\cite{seelig77c}. 
The absolute values of order parameters from the effective dipolar coupling $d_\mathrm{CH}$ ($^1$H-$^{13}$C NMR) are calculated using the equation
$|S_{{\rm CH}}|=\frac{4\pi\langle r_\mathrm{CH}^3 \rangle}{\hbar \mu_0 \gamma_h \gamma_c} d_\mathrm{CH}$, where
values between 20.2--22.7~kHz are used for $\frac{4\pi\langle r_\mathrm{CH}^3 \rangle}{\hbar \mu_0 \gamma_h \gamma_c}$,
depending on the original authors~\cite{hong95a,gross97,dvinskikh05a,ferreira13}.
The effective dipolar coupling $d_\mathrm{CH}$ is related to the measured dipolar splitting $\Delta \nu_\mathrm{CH}$ 
through a scaling factor that depends on the pulse sequence used in the $^1$H-$^{13}$C NMR experiment~\cite{hong95a,gross97,dvinskikh05a,ferreira13}.
It is important to note that the order parameters measured with different techniques based on different physical interactions are in good agreement
with each other (see Results and Discussion), indicating very high quantitative accuracy of the measurements.
For a more detailed discussion, see Ref.~\citenum{accuracyPOST}.

@@

To get a general idea of the quality of the glycerol backbone and choline headgroup structures in different models, we calculated 
the order parameters for these parts from thirteen different lipid models (Table~\ref{systems}) and 
plotted the results together with experimental values in Fig.~\ref{HGorderparameters}.
Two criteria were used to judge the quality of the model: there must not be significant  {\bf forking} in the $\alpha$ and $\beta$ carbons,
there must be only moderate forking in the g$_3$ carbon and there must be significant forking in the g$_1$ carbon, the {\bf magnitude}
should be preferably inside to the subjective sweet spots determined from experiments (blue shaded regions in Fig.~\ref{HGorderparameters}).
The results for each force field with respect to the above criteria are summarized in Figure~\ref{FullHydrationComparisonTable}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{../DATAreportediINblog/comparisonTable.pdf}
  \newline
  \caption{\label{FullHydrationComparisonTable}
Rough subjective ranking of force fields based on Fig.~\ref{HGorderparameters}.
%
Here "M" indicates a magnitude problem, "F" a forking problem;
letter size increases with problem severity.
%
Color scheme:
"within experimental error" (dark green),
"almost within experimental error" (light green),
"clear deviation from experiments" (light red), and
"major deviation from experiments" (dark red).
%
The $\Sigma$-column shows the total deviation of the force field,
when individual carbons are given weights of 0 (matches experiment), 1, 2, and 4 (major deviation).
%
For full details of the assessment, see Supporting Information.
} 
\end{figure}

None of the studied force fields fulfils these criteria completely, however CHARMM36 is close. 
This is not surprising since the dihedral potentials in this model are tuned to reproduce these 
parameters better against experiments~\cite{klauda10}.
The next models in the list are CHARMM36-UA~\cite{henin08,lee14} and H\"ogberg08~\cite{hogberg08} which is also not surprising since
these models are using CHARMM bonded potentials for glycerol backbone and choline. The fourth and the fifth models in the list, MacRog~\cite{maciejewski14} and
GAFFlipid~\cite{dickson12}, have independently determined dihedral potentials. All the models based on Gromos potentials and Slipids perform less well.
In the present work we subject the CHARMM36, MacRog, GAFFlipid and Berger-POPC-07 to a more careful comparison including the sterospecific labeling  
(Fig.~\ref{HGorderparameters2}), and atomistic level structure and responses to the dehydration and cholesterol content in the following sections.
These models are selected for more detailed studies since they are the best representatives of different dihedral potential parametrization techniques 
(CHARMM36, MacRog, GAFFlipid), and the Berger based models are the most used lipid models in the literature.
